R Notebook

### 2021-12-12 src/main/r/numeric_vs_hpo_for_CRC.Rmd
rerun after propensity score matching

sql <- read_file('../sql/get_labs_for_colorectal_cancer_control.sql')
df_control_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'labs_colorectal_cancer_sampled_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = TRUE)
```

@@

df_labs_and_outcomes_case_control %>% sample_n(10)
```
```{r max.output=-1, message=FALSE}
require(furrr)
plan(multisession, workers = 4)

crc_separate_controls_analysis <- df_labs_and_outcomes_case_control %>%
  group_by(loinc) %>%
  nest() %>%
  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction)) %>%
  select(-data) %>%
  unnest(result)

@@

>>>>

sql <- read_file('../sql/get_labs_for_colorectal_cancer_control.sql')
df_control_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'labs_colorectal_cancer_sampled_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)
```

@@

df_labs_and_outcomes_case_control %>% sample_n(10)
```

```{r max.output=-1, message=FALSE}
require(furrr)
plan(multisession, workers=8)
start_time <- Sys.time()

crc_separate_controls_analysis <- df_labs_and_outcomes_case_control %>%
  group_by(loinc) %>%
  nest() %>%
  ungroup() %>%
  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction, .options = furrr_options(seed = T))) %>%
  select(-data) %>%
  unnest(result)

end_time <- Sys.time()

print(end_time - start_time)

@@

# Age and race matching
The above analysis just sampled the control cohort randomly. In this section, we will determine whether we need to match sex, age and race.

```{r}
## because of our processing, cases without desired labs are not counted here
df_labs_and_outcomes_case_control %>% select(medical_record_number, outcome) %>%
  distinct() %>%
  left_join(bind_rows(df_case_demographics, df_control_demographics), by = 'medical_record_number') %>%
  group_by(outcome, race) %>%
  summarise(n = n())

df_demographics_with_outcomes <- bind_rows(df_case_demographics %>% select(-min_diag_date) %>% mutate(outcome = 'abnormal'),
          df_control_demographics %>% mutate(outcome = 'normal')) %>% mutate(age = as.duration(interval(date_of_birth, today())) / dyears())

df_demographics_with_outcomes <- df_demographics_with_outcomes %>%
  filter(gender %in% c('Male', 'Female')) %>%
  filter_all(all_vars(!is.na(.))) %>% 
  mutate(outcome = as.factor(outcome), 
         gender = as.factor(gender),
         race = as.factor(race)) %>% 
  mutate(outcome = relevel(outcome, 'normal'),
         gender = relevel(gender, 'Male'),
         race = relevel(race, 'White'))

## determine if gender and race are associated with outcome: CRC or no CRC
ncase <- nrow(df_demographics_with_outcomes[df_demographics_with_outcomes$outcome == 'abnormal',])
model <- glm(formula = outcome ~ gender + race + age, data = df_demographics_with_outcomes %>% filter(outcome == 'abnormal') %>% union_all(df_demographics_with_outcomes %>% filter(outcome == 'normal') %>% sample_n(ncase)),
                    family = binomial)

anova(model, test = 'Chisq')
```
```{r}
prediction <- predict(model, newdata = df_demographics_with_outcomes , type = 'response', na.action = na.pass)

df_demographics_with_outcomes$propensity_score <- prediction

ggplot(df_demographics_with_outcomes, aes(x = outcome, y = propensity_score)) +
  geom_violin()
```
```{r}
require(MatchIt)

# it is really slow. have to kill it before finishing
# match_outcome <- matchit(outcome ~ gender + race + age, method = "nearest", data = df_demographics_with_outcomes %>% select(-propensity_score))

# try this stackoverflow method
# https://stackoverflow.com/questions/29672088/speedup-matchit
# still slow, but works
df_demographics_with_outcomes %>% colnames()
m.out <- matchit(outcome ~ medical_record_number, data = df_demographics_with_outcomes, method = "nearest", distance = df_demographics_with_outcomes$propensity_score)
save(m.out, file='m.out.obj')
matched_unique_ids_etc <- match.data(m.out, data = df_demographics_with_outcomes)

matched_unique_ids_etc %>% group_by(outcome) %>% summarise(n =n())
ggplot(matched_unique_ids_etc, aes(x = outcome, y = propensity_score)) +
  geom_violin()

ggplot(matched_unique_ids_etc, aes(x = outcome, y = age)) +
  geom_violin()
```
We need to put back the identifiers of controls to the database in order to selectively retrieve their labs. 
```{r}
dbExecute(dsca, "create table if not exists matched_control_cohort (medical_record_number VARCHAR)")
dbExecute(dsca, "truncate table matched_control_cohort")

insert_mrns <- function(mrns_str, dbconn = dsca){
  sql <- glue('insert into matched_control_cohort values {mrns_str};')
  message(sql)
  dbExecute(dbconn, sql)
}

matched_unique_ids_etc %>% filter(outcome == 'normal') %>% 
  rowwise() %>%
  mutate(batch_number = sample(seq(10), size = 1, replace = TRUE)) %>%
  ungroup() %>%
  group_by(batch_number) %>%
  summarise(sql_query = str_c(str_c("('", medical_record_number, "')"), collapse = ",")) %>% 
  ungroup() %>%
  mutate(sql_query = walk(sql_query, insert_mrns, dbconn = dsca))

# retrieve labs for the matched control cohort
sql <- read_file('../sql/get_labs_for_colorectal_cancer_matched_control.sql')
df_matched_control_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'labs_colorectal_cancer_matched_controls.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = TRUE)

# drop the matched control cohort
# persist it for a while for now
# n <- dbExecute(dsca, "DROP TABLE IF EXISTS matched_control_cohort")
# assertthat::are_equal(n, 0)
```

Rerun the analysis with the matched controls.

# labs for controls
df_labs_and_outcomes_matched_controls <- df_matched_control_labs %>% filter(!(loinc %in% c('21000-5', '32207-3'))) %>%
  as_tibble() %>%
  #filter(medical_record_number == '7007952')  %>%
  #mutate(lab_date = as.Date(lab_date)) %>%
  arrange(medical_record_number, loinc, desc(lab_date)) %>%
  group_by(medical_record_number, loinc) %>%
  mutate(sequence = row_number()) %>%
  ungroup() %>%
  filter(sequence == 1) %>%
  mutate(outcome = 'normal')

nrow(df_labs_and_outcomes_cases)
nrow(df_labs_and_outcomes_matched_controls)

df_labs_and_outcomes_case_matched_control <- bind_rows(
  df_labs_and_outcomes_cases %>% select(medical_record_number, lab_date, loinc, local_test_code, lab_result_numeric_value, unit_of_measure, normal_range, interpretation, hpotermid, isnegated, outcome),
  df_labs_and_outcomes_matched_controls %>% select(medical_record_number, lab_date, loinc, local_test_code, lab_result_numeric_value, unit_of_measure, normal_range, interpretation, hpotermid, isnegated, outcome)
)

rm(df_labs_and_outcomes_cases, df_labs_and_outcomes_matched_controls)

df_labs_and_outcomes_case_matched_control %>% sample_n(10)
```

```{r max.output=-1, message=FALSE}
require(furrr)
plan(multisession, workers=4)
start_time <- Sys.time()

crc_case_matched_controls_analysis <- df_labs_and_outcomes_case_matched_control %>%
  group_by(loinc) %>%
  nest() %>%
  ungroup() %>%
  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction, .options = furrr_options(seed = T))) %>%
  select(-data) %>%
  unnest(result)

end_time <- Sys.time()

print(end_time - start_time)

crc_case_matched_controls_analysis
```
